---
title: "Learning Bayesian Networks from gene expression data"
author: Modestas Filipavičius
date: February 22, 2018
output: pdf_document

---

Here we use R package deal1 to analyse yeast cell-cycle regulator genes and learn optimal Bayesian Network.

# Install the packages and load and preprocess the data.
YeastCC data: This data package contains an ExpressionSet instance for the yeast cell cycle microarray experiment. The dataset contains gene expression measures (log-ratios, with Cy3-labeled common reference) for 6,178 yeast genes in 77 conditions.
```{r, message=F, warning=F}
# Install deal and load yeastCC data
# install.packages("deal")
library(deal)
#source("https://bioconductor.org/biocLite.R")
#biocLite("yeastCC")
library(yeastCC)

# Load the Yeast cell cycle regulator dataset
yeastCC = yeastCC
# First 3 genes are: [1] "YAL001C"   "YAL002W"   "YAL003W"
# Last genes are:    [1] "YPR202W" "YPR203W" "YPR204W"

# 77 conditions for each gene are measured in log-ratios: 
#          cln3.1 cln3.2 clb2.2 clb2.1 alpha0 alpha7 alpha14 alpha21 alpha28 alpha35
# YAL001C   0.15     NA  -0.22   0.07  -0.15  -0.15   -0.21    0.17   -0.42   -0.44
# length(exprs(yeastCC[1]))
```


## Turn data object yeastCC into an expression matrix (function exprs) with genes as columns and experiments as rows.
```{r}
expr_df = as.data.frame(exprs(yeastCC))
expr_mt = t(as.matrix(expr_df))
expr_mt[1:5, 1:5]
```


## Restrict the genes to cell cycle regulators.
The ORF names for the 800 cell cycle regulated genes are stored in orf800.
```{r}
# Create a mask with yeast gene indeces that correspond to 800 cell regulator genes
mask = which(colnames(expr_mt) %in% orf800)

# Subset cell regulators
expr_cc_mt = expr_mt[ , mask]
expr_cc_mt[1:5, 1:5]

```


## Replace missing data (=NA) with each gene’s median.
```{r}
replace_na_with_col_medians = function(x) {
    x = replace(x, is.na(x), median(x, na.rm = T))
    return(x)
}

cat("Number of NA's before cleanup is:", length(which(is.na(expr_cc_mt) == TRUE)))
expr_cc_mt = apply(expr_cc_mt, 2, replace_na_with_col_medians)
cat("Number of NA's after cleanup is:", length(which(is.na(expr_cc_mt) == TRUE)))
# # Check if correct w/ 1st and 2nd genes
# which(is.na(expr_cc_mt[, 2]))           # cdc28.140 idx61; cdc28.110 idx58
# median(expr_cc_mt[, 2], na.rm = T)      # -0.5  -0.045

```


## Calculate the range between the 25th and 75th quantile for each gene and keep genes with a value > 1.6.
The interquartile range of an observation variable is the difference of its upper and lower quartiles. It is a measure of how far apart the middle portion of data spreads in value.

```{r}
# Calculate IQRs for every gene/column, select IQRs with >1.6 and subset the data with them
iqrs = apply(expr_cc_mt, 2, IQR)
mask = which(iqrs > 1.6)
expr16 = as.data.frame(expr_cc_mt[, mask])
expr16[1:5,1:5]

```

## Add a dummy gene with expression 1 over all experiments (deal needs at least one discrete variable).
```{r}
YOLO = as.factor(rep(1, length(expr16[, 1]))) # must be a factor because discrete variable expected in deal
expr16 = cbind(expr16, YOLO)
cat("Class of dummy discrete variable should be a factor:", class(expr16[1, 11]))

```




# Learn the optimal network

## Create an empty network prior.
```{r}
bn_expr16 = network(expr16) # to visually inspect local probs: inspectprob=TRUE
# nodes(bn_expr16) # YOLO node should be discrete!
plot(bn_expr16)
```

## What are the local probability distributions of the genes?

```{r}
localprob(bn_expr16)
```


## Ban all outgoing edges for the dummy node (use function banlist). 
```{r}
# build a table with edges between 11 and other nodes
mybanlist = matrix(c(1,2,3,4,5,6,7,8,9,10, 
                     11,11,11,11,11,11,11,11,11,11), 
                   ncol=2)

banlist(bn_expr16) =  mybanlist
plot(bn_expr16)
```

## Create a joint prior with imaginary sample size 5, N=5
```{r}
bn_expr16.prior = jointprior(bn_expr16, N=5)
head(bn_expr16.prior)
```

## Learn the initial network (use getnetwork to turn the result into an actual network).
```{r}
# master prior
bn_expr16.m_prior = learn(bn_expr16, expr16, bn_expr16.prior)

# initial network
bn_expr16.initial = getnetwork(bn_expr16.m_prior)
cat("Initial network score is:", bn_expr16.initial$score)
```
```{r fig.height = 5, fig.width = 5, fig.align = "center"}
plot(bn_expr16.initial, showban = FALSE)
```


## Search for the optimal network and plot it
```{r include=FALSE}
bn_expr16.optimal = autosearch(bn_expr16.initial, expr16, bn_expr16.prior, showban = F, trace = F, timetrace = F)
bn_expr16.optimal = getnetwork(bn_expr16.optimal)
```
```{r fig.height = 5, fig.width = 5, fig.align = "center"}
plot(bn_expr16.optimal, showban = F)
```

## Optimal network score is: 
```{r}
bn_expr16.optimal$score
```
